\documentclass[aspectratio=169,11pt]{beamer}

% ============================================================
% SISMID 2026 -- Day 1, 3:30 session
% Using AI agents for data scraping (S. Yang + C. Djorno)
% General agent-scraping recipe -> a menu of streams (Wikipedia,
% wastewater, mobility, news) -> elaborate agent-driven scraping
% -> when scraping is blocked: proxy / VPN -> Plan B.
% (Proxy/VPN discussion lives HERE, not in the 9:45 first encounter.)
% Compile: pdflatex ai-agents-data-scraping.tex  (run twice)
% ============================================================

\IfFileExists{beamerthememetropolis.sty}{
  \usetheme{metropolis}
  \metroset{progressbar=frametitle, sectionpage=progressbar, numbering=fraction}
}{
  \usetheme{Boadilla}
  \usecolortheme{seahorse}
  \setbeamertemplate{navigation symbols}{}
}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, fit, backgrounds, calc}
\usepackage{xcolor}
\usepackage{listings}

\definecolor{AccentA}{HTML}{2A6F97}
\definecolor{AccentB}{HTML}{C1666B}
\definecolor{AccentC}{HTML}{3B8A5B}
\setbeamercolor{frametitle}{bg=AccentA}
\setbeamercolor{progress bar}{fg=AccentB}

\definecolor{skbg}{rgb}{0.96,0.96,0.94}
\lstdefinestyle{prompt}{
  basicstyle=\ttfamily\scriptsize,
  backgroundcolor=\color{skbg},
  showstringspaces=false, breaklines=true,
  frame=single, framesep=4pt, columns=fullflexible, numbers=none,
}
\lstset{style=prompt}

\tikzset{
  cbox/.style={draw=AccentA, very thick, rounded corners, align=center,
    minimum height=1.0cm, inner sep=6pt, font=\small, fill=AccentA!6},
  sbox/.style={draw=AccentA, thick, rounded corners, align=center,
    minimum height=0.85cm, inner sep=4pt, font=\scriptsize, fill=AccentA!6},
  rbox/.style={draw=AccentB, thick, rounded corners, align=center,
    minimum height=0.85cm, inner sep=4pt, font=\scriptsize, fill=AccentB!8},
  gbox/.style={draw=AccentC, thick, rounded corners, align=center,
    minimum height=0.85cm, inner sep=4pt, font=\scriptsize, fill=AccentC!8},
  flow/.style={-{Stealth[length=3mm]}, very thick, draw=AccentB},
  flowq/.style={-{Stealth[length=2.4mm]}, thick, draw=AccentA!70},
}

\title{Using AI Agents for Data Scraping}
\subtitle{Google Trends for your own topic, and getting everyone started}
\author{Shihao Yang \quad\textbullet\quad Candice Djorno}
\institute{SISMID 2026 \textbullet\ Day 1, 3:30}
\date{July 20, 2026}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{From one topic to your topic}
  This morning you scraped Google Trends for \textbf{one} topic (dengue in Mexico). This
  session gets \textbf{everyone} comfortable scraping Google Trends for \textbf{their own}
  disease and place, before we add other streams tomorrow.
  \begin{block}{The goal this session}
    A reliable, repeatable way to pull Google Trends for whatever you care about, with the
    agent doing the typing and you keeping the judgment.
  \end{block}
  \vspace{0.3em}
  \begin{itemize}
    \item The agent is the \textbf{coding tool woven through the whole course}; today we make
          it second nature on the stream you already know.
    \item \textbf{Plan B} pre-filled notebooks exist, so preferring not to drive an agent
          never blocks you.
  \end{itemize}
\end{frame}

\begin{frame}{Roadmap}
  \tableofcontents
\end{frame}

% ===============================================================
\section{The general recipe}

\begin{frame}{Every stream, the same five steps}
  \begin{center}
  \resizebox{0.96\textwidth}{!}{%
  \begin{tikzpicture}[node distance=0.9cm]
    \node[sbox] (f) {\textbf{fetch}\\call the API\\(or agent browser)};
    \node[sbox, right=of f] (t) {\textbf{tidy}\\date + value\\columns};
    \node[sbox, right=of t] (p) {\textbf{plot}\\look before\\trusting};
    \node[rbox, right=of p] (s) {\textbf{sanity-check}\\range, gaps,\\reproducible?};
    \node[gbox, right=of s] (v) {\textbf{save}\\tidy CSV};
    \draw[flow] (f) -- (t); \draw[flow] (t) -- (p); \draw[flow] (p) -- (s); \draw[flow] (s) -- (v);
  \end{tikzpicture}}
  \end{center}
  \vspace{0.4em}
  \begin{center}
    \itshape You learned this on Google Trends this morning. It does not change from stream
    to stream. Only the fetch step differs, and the agent handles that.
  \end{center}
\end{frame}

\begin{frame}[fragile]{Anatomy of a good scrape prompt}
  Four ingredients: the \textbf{outcome}, the \textbf{source}, the \textbf{shape}, and the
  \textbf{check}.
  \begin{block}{Template}
  \begin{lstlisting}
Pull <what series> for <place/topic> over <time window>
from <source / API>. Return a tidy DataFrame: a date
column plus one column per series. Then plot it and tell
me <the check: date range, missing values, does it look
seasonal?>.
  \end{lstlisting}
  \end{block}
  \begin{center}
    \itshape The check is not optional. It is what keeps the agent honest and what teaches
    you to read the output.
  \end{center}
\end{frame}

\begin{frame}[fragile]{Scrape it for your own topic}
  The whole point: swap in \emph{your} disease and place. One sentence changes.
  \begin{block}{Prompt (edit the two bold parts)}
  \begin{lstlisting}
Using pytrends, pull Google Trends for my topic:
terms about *RSV* in *the United States* (geo=US),
past 5 years. Tidy date + one column per term, retry on
429, plot, and print the last date.
  \end{lstlisting}
  \end{block}
  \begin{itemize}
    \item Everyone in the room can pull a \emph{different} disease and place at the same
          time. No one pre-stages a CSV for each.
    \item This is the ``you no longer need a HealthMap'' idea: the agent assembles a signal
          for \emph{your} county and \emph{your} disease on the spot.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Term vs topic (and building a topic from phrases)}
  Google Trends can give you two different things for ``flu'':
  \begin{itemize}
    \item a \textbf{term} (the literal string), which misses synonyms, spellings, and other
          languages, and shows more zeros; or
    \item a \textbf{topic} (Google's aggregated Knowledge Graph entity), which folds all of
          those together, with more volume and fewer zeros.
  \end{itemize}
  Two ways to reach a topic-like signal with the agent:
  \begin{block}{Prompt}
  \begin{lstlisting}
Pull three things for my disease and compare: (1) the
individual terms, (2) one combined additive query joining
the phrases with " + " (a topic-like OR), and (3) the real
Google Trends topic entity (resolve it with pytrends
suggestions and pull by its mid). Which has the fewest
zeros and the strongest seasonal signal?
  \end{lstlisting}
  \end{block}
\end{frame}

\begin{frame}{Verify what the agent produced (still your job)}
  Fast is not the same as correct. Every scraped stream earns a few cheap checks:
  \begin{itemize}
    \item \textbf{Range and gaps:} does the span match what you asked? Missing weeks become
          wrong models.
    \item \textbf{Scale sanity:} plausible values, not all zeros or a flat line from a
          failed pull.
    \item \textbf{Reproducibility:} pull twice. Google Trends is noisy, so two pulls can
          differ; that is the sampling issue from this afternoon.
    \item \textbf{Face validity:} do the peaks land in the right season?
  \end{itemize}
  \vspace{0.3em}
  \begin{alertblock}{The habit}
    One plot and one sanity question per stream, before it enters any model.
  \end{alertblock}
\end{frame}

% ===============================================================
\section{Making the scrape robust}

\begin{frame}{Scraping vs an API: use the key}
  \texttt{pytrends} scrapes an \emph{unofficial} endpoint. It works, and it teaches the
  gotchas, but it is not what you would run in production.
  \begin{center}
  \footnotesize
  \begin{tabular}{lll}
    \hline
     & \textbf{pytrends (scraping)} & \textbf{Trends API (\texttt{getGraph})} \\
    \hline
    Access      & unofficial endpoint      & documented API, needs a key \\
    Failures    & HTTP 429 roulette        & none in normal use \\
    Values      & \textbf{resampled each pull} & \textbf{identical every pull} \\
    Signal      & normalized 0--100        & \textbf{the same} normalized 0--100 \\
    \hline
  \end{tabular}
  \end{center}
  \vspace{0.3em}
  \begin{alertblock}{Same signal, no roulette}
    The API is a \textbf{drop-in replacement}: the same 0--100 index, the same weekly or
    monthly resolution chosen by window length. You get a class key today.
  \end{alertblock}
\end{frame}

\begin{frame}[fragile]{Using the class key}
  \begin{block}{Set it once, before you start Jupyter}
  \begin{lstlisting}
export GT_API='the key handed out in class'
  \end{lstlisting}
  \end{block}
  One call, no retry loop:
  \begin{lstlisting}
GET https://www.googleapis.com/trends/v1beta/graph
    ?key=$GT_API
    &terms=dengue&terms=/m/0cycc      # terms OR topic ids
    &restrictions.geo=MX              # US, MX, US-GA ...
    &restrictions.startDate=2021-07   # YYYY-MM
    &restrictions.endDate=2026-07
  \end{lstlisting}
  \begin{itemize}
    \item \textbf{Quota: 1,000 requests/day}, and \textbf{health topics only}. Please do not
          burn it on finance data.
    \item \textbf{Never commit the key.} It lives in an environment variable, not in a
          notebook.
  \end{itemize}
\end{frame}

\begin{frame}{Term vs topic: the language lesson}
  The API takes a literal \textbf{term} or a Knowledge Graph \textbf{topic id}
  (\texttt{/m/0cycc} = Influenza), which folds together every spelling, synonym and
  \emph{language} for the same concept.
  \begin{center}
  \footnotesize
  \begin{tabular}{lcccc}
    \hline
    \textbf{Country} & \multicolumn{2}{c}{\textbf{term ``flu''}} & \multicolumn{2}{c}{\textbf{topic /m/0cycc}} \\
                     & max & mean & max & mean \\
    \hline
    United States & 100 & --   & 100 & -- \\
    France        & 3   & 0.9  & \textbf{100} & \textbf{16.3} \\
    Italy         & 4   & 1.6  & \textbf{100} & \textbf{17.7} \\
    Mexico        & 4   & 1.7  & \textbf{100} & \textbf{37.1} \\
    \hline
  \end{tabular}
  \end{center}
  \vspace{0.2em}
  \begin{center}
    \itshape An English term is nearly silent abroad. The topic sees \emph{grippe},
    \emph{influenza} and \emph{gripe} alike, which is why cross-country work uses topics.
  \end{center}
\end{frame}

\begin{frame}{Let the agent run the whole loop}
  A single, richer prompt can chain the steps and leave you an artifact to inspect:
  \begin{center}
  \resizebox{0.95\textwidth}{!}{%
  \begin{tikzpicture}[node distance=0.8cm]
    \node[sbox] (a) {find the\\API / docs};
    \node[sbox, right=of a] (b) {write the\\scraper};
    \node[sbox, right=of b] (c) {run + fix\\errors};
    \node[rbox, right=of c] (d) {sanity-\\check};
    \node[gbox, right=of d] (e) {cache a\\snapshot};
    \draw[flow] (a) -- (b); \draw[flow] (b) -- (c); \draw[flow] (c) -- (d); \draw[flow] (d) -- (e);
  \end{tikzpicture}}
  \end{center}
  \vspace{0.4em}
  \begin{center}
    \itshape You stay in the loop at the sanity-check. The agent does the typing and the
    debugging around it.
  \end{center}
\end{frame}

% ===============================================================
\section{When scraping is blocked: proxy and VPN}

\begin{frame}{Getting blocked is a normal part of scraping}
  Pull hard enough, from the wrong kind of address, and a server pushes back with a
  \textbf{429} or an outright block. Three tiers, in order:
  \begin{itemize}
    \item \textbf{First: retry.} Most 429s clear on a wait-and-retry. We tested Google Trends
          from a Codespace on an \textbf{Azure} IP: 429, wait, then success. Usually enough.
    \item \textbf{Backup: a proxy or VPN.} If retries keep failing (a whole room at once, or a
          venue whose IP is blocked), route through a machine with an ordinary IP.
    \item \textbf{Last resort: the cache.} Every stream ships a cached snapshot, so no one is
          ever stuck.
  \end{itemize}
  \vspace{0.2em}
  \begin{center}
    \itshape We would rather route around a block than accept it. The cache is the floor, not
    the plan.
  \end{center}
\end{frame}

\begin{frame}{What is a proxy or a VPN?}
  Every request your computer sends carries a ``return address'' (its IP). When a site blocks
  that address, you route the request through \emph{another} machine, so the site sees
  \emph{its} address instead of yours.
  \begin{itemize}
    \item \textbf{Proxy:} reroute just one app (here, the data download).
    \item \textbf{VPN:} reroute \emph{all} of a machine's traffic the same way.
  \end{itemize}
  A spare laptop on ordinary home internet becomes a normal-looking doorway for our cloud
  Codespaces, whose Azure address some sites throttle.
  \vspace{0.2em}
  \begin{center}
    \itshape Same trick a traveler uses to watch home TV abroad. Nothing exotic.
  \end{center}
\end{frame}

\begin{frame}[fragile]{You do not set this up by hand: ask an agent}
  Standing up a proxy used to be a sysadmin chore. Now you describe the goal in plain words
  and let a coding agent handle the details:
  \begin{block}{Prompt (paste into the agent on your spare machine)}
  \begin{lstlisting}
I'm teaching a class and a website is blocking our shared
cloud computers from downloading data. I have a spare
laptop on normal home internet. Set it up so my students
can send their downloads through this machine, and tell me
in plain steps what to run and exactly what each student
types. Keep it simple and secure.
  \end{lstlisting}
  \end{block}
  \begin{center}
    \itshape The agent picks the tool, sets a password, and hands you the student line.
    We already built ours; here is the proof.
  \end{center}
\end{frame}

\begin{frame}[fragile]{We built one, and it works}
  We stood up exactly this: a spare machine on ordinary home broadband running a small,
  password-protected proxy (tinyproxy), reachable from anywhere via a tunnel (ngrok).
  \begin{block}{Proof, from a real Codespace routed through it}
  \begin{lstlisting}
egress via proxy: a residential IP   (home broadband, NOT Azure)
homepage: 200   token: 200   data: 200
SUCCESS: 92 weekly points; last point = this week
VERDICT: Google Trends scraping WORKS via the proxy
  \end{lstlisting}
  \end{block}
  \begin{center}
    \itshape A student turns it on with one line:
    \texttt{export HTTPS\_PROXY=http://user:pass@host:port}
  \end{center}
\end{frame}

% ===============================================================
\section{Plan B and wrap-up}

\begin{frame}{Plan B is a first-class path}
  Not everyone wants to drive an agent, and that is fine.
  \begin{itemize}
    \item Every stream has a \textbf{pre-filled notebook} that is a captured example of what
          the agent produces from these prompts. Run it top to bottom.
    \item Every stream has a \textbf{cached snapshot}, so a failed or blocked live pull is a
          non-event.
    \item Pair up: an experienced neighbor plus the agent is the fastest way to learn what
          good output looks like.
  \end{itemize}
  \vspace{0.3em}
  \begin{center}
    \itshape Lane A (agent) and Lane B (notebook) reach the same tidy CSV. Switch any time.
  \end{center}
\end{frame}

\begin{frame}{What you can now do}
  \begin{itemize}
    \item Scrape \textbf{Google Trends} for \textbf{your own} disease and place with one
          prompt, and check the result yourself.
    \item Make it robust: retry on 429, keep the sampling gotchas in mind.
    \item Get \textbf{around a block} with retry, then a proxy or VPN, with the cache as a
          floor.
    \item Fall back to Plan B notebooks whenever you prefer.
  \end{itemize}
  \vfill
  \begin{center}
    \itshape The agent is your data-collection layer for the rest of the week.
  \end{center}
\end{frame}

\begin{frame}{Looking ahead: Day 2}
  \begin{itemize}
    \item \textbf{Tomorrow morning:} data beyond Google Trends, Wikipedia pageviews and
          wastewater in depth, scraped the same way you did today.
    \item \textbf{Then:} mobility and news, and folding every stream into ARGO and the
          capstone.
  \end{itemize}
  \vfill
  \begin{center}
    \itshape You have the method. Day 2 is just more streams through the same loop.
  \end{center}
\end{frame}

\begin{frame}[standout]
  Questions?\\[0.4em]
  \small Repo: \texttt{github.com/Shihao-Yang/sismid2026}
\end{frame}

\end{document}
